Assessing Racial Bias in Arrest and Booking Practices in Toronto: Insights and Implications from 2020-2021

### 2024-12-03 paper/paper.qmd
Rewrote sections of appendix

---
title: "Analyzing Arrest Patterns and Racial Representation in Toronto"
author: "Akshat Aneja"
abstract: This paper analyzes arrest data from the Toronto Police Service for the years 2020 and 2021, focusing on the distribution of arrests by ethnic identity, age, and booking status to assess whether systemic bias exists in policing practices. The analysis reveals that, while arrests appear to be driven by objective factors, disparities related to socio-economic status may still exist. These findings provide important insights that can help improve transparency and accountability in policing, contributing to more equitable law enforcement practices.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Akshat211202/bias-in-arrests-toronto"

@@

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "# arrests by Ethnicity in Year 2020 and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig1

@@

# Create a plot
ggplot(table1_long, aes(x = perceived_race, y = value, fill = year)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(x = "Ethnicity",
       y = "# Arrests",
       fill = "Year") +
  theme_minimal() +
  scale_fill_discrete(labels = c("2020", "2021")) +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

@@

@@

(@fig-ppc) shows posterior predictive check was performed . @fig-ppc illustrates a comparison between posterior distribution simulations and the actual probability of experiencing a custody arrest

@@

In addition, using code adapted from @tellingstorieswithdata, a comparison of the prior and posterior was performed (@fig-comparison). @fig-comparison depicts the amounts by which the estimates are modified after considering data [@tellingstorieswithdata].

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-comparison
#| fig-cap: "Comparison of the prior and posterior"

# The code below was adapted based on Alexander (2023).
# 
# posterior_vs_prior(model) +
#   theme_minimal() +
#   scale_color_brewer(palette = "Set1") +
#   theme(legend.position = "none") +
#   coord_flip()
```

@@

This section assesses whether the `rstanarm` [@citerstanarm] Markov chain Monte Carlo (MCMC) algorithm experienced problems [@tellingstorieswithdata].

Using code adapted from @tellingstorieswithdata, a trace plot was generated (@fig-trace). @fig-trace illustrates horizontal lines that are seemingly erratic with chain overlap, indicating nothing unusual (based on @tellingstorieswithdata).

@@

Moreover, using code adapted from @tellingstorieswithdata, a Rhat plot was created (@fig-rhat). @fig-rhat depicts values near 1 and less than or equal to 1.1, indicating no issues (based on @tellingstorieswithdata).

@@

# The code below was adapted from Alexander (2023).

@@

Using code adapted from @tellingstorieswithdata, @fig-credibility-intervals-all was generated. @fig-credibility-intervals-all illustrates 90\% credibility intervals of all of the predictors of undergoing a custody arrest based on the model from this report.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-credibility-intervals-all
#| fig-cap: "Credibility intervals of all of the custody arrest predictors"

# The code below was adapted from Alexander (2023).

@@

@fig-credibility-intervals-several depicts 90\% credibility intervals of several of the predictors of undergoing a custody arrest. In particular, the American Indian/Alaskan Native and Hawaiian/Pacific Islander predictors were omitted for visual clarity (as compared to @fig-credibility-intervals-all).

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-credibility-intervals-several
#| fig-cap: "Credibility intervals of several of the custody arrest predictors"

# Portions of the code below were adapted from Alexander (2023).

>>>>

---
title: "Assessing Racial Bias in Arrest and Booking Practices in Toronto: Insights and Implications from 2020-2021"
subtitle: "Black and Indigenous Overrepresentation in Arrests with Stable Booking Probabilities Across Ethnic Groups"
author: "Akshat Aneja"
abstract: This study examines arrest and booking data from the Toronto Police Service for 2020 and 2021 to explore potential racial disparities in policing practices. While Black and Indigenous individuals were disproportionately represented in arrests, the likelihood of being formally booked after arrest was similar across ethnic groups. These findings suggest that disparities may occur earlier in the law enforcement process rather than at the booking stage. By highlighting these patterns, the study underscores the need for targeted reforms to ensure equity and transparency in policing practices.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Akshat211202/bias-in-arrests-toronto"

@@

The study seeks to estimate the relative risk (or odds) of arrest for individuals belonging to specific racial groups compared to a baseline racial group, adjusting for factors such as socioeconomic status, neighborhood characteristics, and crime type. This parameter reflects the proportion of observed disparities that can be attributed to racial representation in the city's population.

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Number of arrests by Ethnicity in Year 2020 and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig1

@@

# Create a plot
ggplot(table1_long, aes(x = perceived_race, y = value, fill = year)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(x = "Ethnicity",
       y = "Number of Arrests",
       fill = "Year") +
  theme_minimal() +
  scale_fill_discrete(labels = c("2020", "2021")) +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

@@

\appendix

@@

[@fig-ppc] shows a posterior predictive check was performed. It illustrates a comparison between posterior distribution simulations and the actual probability of experiencing a custody arrest

@@

@@

[@fig-trace] shows well-mixed and overlapping chains, indicating good convergence of the Bayesian logistic regression model's Markov Chain Monte Carlo sampling process. This validates the reliability of the posterior estimates reported in the paper, supporting the conclusion that booking probabilities were consistent across ethnic groups and years. However, the wide credibility intervals for some coefficients highlight the need for cautious interpretation, especially for smaller ethnic groups, due to limited data and high variability.

@@

[@fig-rhat] demonstrates that all Rhat values are approximately 1.0, which indicates that the Markov Chain Monte Carlo (MCMC) chains have converged well. This suggests that the Bayesian logistic regression model used in the paper is reliable, as the convergence ensures robust and consistent parameter estimation. These results support the validity of the study's findings on arrest and booking probabilities, providing confidence in the statistical inferences drawn from the model.

@@

@@

The credibility intervals in [@fig-credibility-intervals-all] show that most coefficients, including race and year, have intervals crossing zero, indicating no statistically significant effects of these predictors on booking probabilities. This aligns with the paper's findings that booking probabilities were consistent across ethnic groups and years, suggesting limited evidence of systemic bias at the booking stage. However, the wide intervals, particularly for smaller ethnic groups like Indigenous and East/Southeast Asian, reflect data limitations and highlight the need for caution in drawing definitive conclusions.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-credibility-intervals-all
#| fig-cap: "Credibility Intervals for All Predictors in Booking Analysis"

@@

The credibility intervals in [@fig-credibility-intervals-several]  focus on a subset of predictors and show that all intervals overlap with zero, indicating no statistically significant effects of these predictors on booking probabilities. This reinforces the paper's conclusion that booking probabilities are consistent across ethnic groups and between years, with no strong evidence of systemic bias at the booking stage. However, the wide intervals for smaller groups like Indigenous and East/Southeast Asian highlight data limitations and the need for caution in interpreting the results.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-credibility-intervals-several
#| fig-cap: "Credibility Intervals for Key Predictors in Booking Analysis"
